% Chapter01.tex
% Copyright (c) 2014 - The Foreseeable Future, zhiayang@gmail.com

\input{Preamble}


\section{Reaction Mechanisms}

\subsection{Bond Breaking}
	\subsubsection{Homolytic Fission}

		Homolytic fission involves the splitting of a single bond, with an equal (\textit{homo}) distribution of the two electrons
		of the aforementioned bond. This results in the formation of radicals, since now both atoms from the bond will have
		unpaired electrons.

		The movement of these single electrons is represented with single-hooked arrows. The arrow starts from the bond line, and points
		towards the target atom receiving the electron.

		\diagram{
			\schemestart[0, 1.5, thick]
			\chemfig{@{carb}C(-[@{bond}:0]@{hyd}H)(-[:90]H)(-[:180]H)(-[:270]H)}
			\arrow{->}
			\chemfig{\lewis{0.,C}(-[:90]H)(-[:180]H)(-[:270]H)} \hspace{0.5cm} \textbf{+} \hspace{0.5cm} \chemfig{\lewis{4.,H}}
			\schemestop

			% draw arrows
			\chemmove{\draw[-{Stealth[left]}, line width=0.4mm, shorten <=1mm, shorten >=1mm](bond).. controls +(90:5mm) and +(120:8mm).. (hyd);}
			\chemmove{\draw[-{Stealth[left]}, line width=0.4mm, shorten <=1mm, shorten >=1mm](bond).. controls +(270:5mm) and +(285:8mm).. (carb);}

		}{The homolytic fission of \ce{CH4}, forming \ce{CH3} and \ce{H} radicals.}

	% end subsubsection

	\subsubsection{Heterolytic Fission}

		As the name implies, heterolytic fission is the opposite; it distributes both electrons of the bond to a single atom, which
		usually results in the formation of ions. The transfer here is represented using a full (double-hooked) arrow.


		\diagram{
			\schemestart[0, 1.5, thick]
			% note: chlorine has to be coloured manually here
			% because somewhere in chemfig it fucks up, you can't name a submolecule.
			\chemfig{@{carb}C(-[@{bond}:0]@{chl}\color{OliveGreen}\chlorine)(-[:90]H)(-[:180]H)(-[:270]H)}
			\arrow{->}
			\chemfig{C|\sps{+}(-[:90]H)(-[:180]H)(-[:270]H)}
			\hspace{0.5cm} \textbf{+} \hspace{0.5cm}
			\chemfig{\lewis{4:,\color{OliveGreen}\chlorine}}
			\schemestop

			% draw arrows
			\chemmove{\draw[-{Stealth}, line width=0.4mm, shorten <=1mm, shorten >=1mm](bond).. controls +(90:5mm) and +(120:8mm).. (chl);}


		}{The heterolytic fission of \ce{CH3\chlorine}, resulting in \ce{CH3+} and \ce{\chlorine-} ions}

	% end subsubsection


\subsection{Bond Forming}
	\subsubsection{Single Electrons}

		Bond forming is essentially the reverse of bond breaking, and hence the same notations apply –– single-hooked arrow for the
		movement of single electrons, and double-hooked arrows for the movement of an electron pair.

		In this case, two single electrons from each atom contribute to the new bond.


		\diagram{
			\schemestart[0, 1.5, thick]
			\chemfig{@{carb}\lewis{1.,C}(-[:90]H)(-[:180]H)(-[:270]H)}
			\hspace{0.5cm} \chemfig{@{plus}\textbf{+}} \hspace{0.5cm}
			\chemfig{@{hyd}\lewis{5.,H}}
			\arrow{->}
			\chemfig{C(-[:0]H)(-[:90]H)(-[:180]H)(-[:270]H)}
			\schemestop

			% draw arrows
			\chemmove{\draw[-{Stealth[left]}, line width=0.4mm, shorten <=2mm, shorten >=1mm](carb).. controls +(45:5mm) and +(120:8mm).. (plus);}
			\chemmove{\draw[-{Stealth[left]}, line width=0.4mm, shorten <=2mm, shorten >=1mm](hyd).. controls +(225:5mm) and +(285:8mm).. (plus);}

		}{The formation of a bond between \ce{CH3} and \ce{H} radicals to form \ce{CH4}.}


	% end subsubsection


	\subsubsection{Electron Pairs}

		Electron pairs usually come from negatively charged radicals or lone pars, although this is not a rule.
		Both electrons come from a single source to form a bond, but this is \textit{not} a dative bond –– it is simply a normal
		bond.

		\diagram{
			\schemestart[0, 1.5, thick]
			\chemfig{@{carb}C|\sps{+}(-[:90]H)(-[:180]H)(-[:270]H)}
			\hspace{0.5cm} \chemfig{\textbf{+}} \hspace{0.5cm}
			\chemfig{@{chl}\lewis{3:,\color{OliveGreen}\ce{\chlorine-}}}
			\arrow{->}
			\chemfig{C(-[:0]!{Cl})(-[:90]H)(-[:180]H)(-[:270]H)}
			\schemestop

			% draw arrows
			\chemmove{\draw[-{Stealth}, line width=0.4mm, shorten <=3mm, shorten >=1mm](chl).. controls +(150:12mm) and +(45:12mm).. (carb);}

		}{The formation of a bond between \ce{CH3} and \ce{\chlorine-} radicals, to form \ce{CH3\chlorine}.}

	% end subsubsection


















% end subsection




% end subsection

