% Chapter 03 - Reaction Kinetics.tex
% Copyright (c) 2014 - 2016, zhiayang@gmail.com
% Licensed under the Apache License Version 2.0.

\pagebreak
\part{Reaction Kinetics}

	\section{Collision Theory}
		The collision theory of reaction is used to represent how individual molecules of reactants actually behave in a given reaction.
		It models reactions as collisions between molecules, and thus there are three conditions that must be satisfied:

		\begin{bulletlist}
			& A collision must occur
			& The collision must occur with the molecules in the correct orientation
			& The molecules must collide with sufficient speed to overcome the activation energy
		\end{bulletlist}

		Furthermore, the rate of a given reaction depends on the number of effective collisions in a given amount of time --- the
		\itl{effective collision frequency}.

		The importance of having the correct orientation can be seen in the electrophilic addition of \ch{HBr} to an alkene; the \deltap{}
		on the \ch{H} atom is attacked by the electron-rich $\pi$-bond, so naturally the \ch{HBr} molecule must be oriented such that the \ch{H}
		end is facing the alkene.

	% end section



	\section{Activation Energy}

		Activation energy represents the minimum amount of energy required before a reaction can take place --- this is usually manifested as
		the kinetic energy of colliding molecules of reactants, using the collision theory model of reaction. For most reactions, the activation
		energy is sufficiently high that the rate of reaction is not ridiculously fast, given that there are an enormous number of collisions
		between molecules at any given time.


		\pagebreak
		\subsection{Energy Profile Diagram}

			Activation energy is often represented in two ways; firstly, in the energy profile diagram as shown below. Note that even though
			the reaction is exothermic (it releases energy), it still requires a minimum amount of activation energy (\ea) to take place. On a side
			note, at the top of the energy \enquote{peak}, a theoretical activated complex or transition state is formed; this cannot be isolated, and
			exists only for a tiny amount of time.


			\graphdiagram[1.0]{
				\begin{endiagram}[scale=2.0,offset=1.5,x-label-text=reaction coordinate,y-label-text=energy]
					\ENcurve[step=1.5]{2,4.5,0}

					\ShowNiveaus[shift=-1.0,length=2.0,niveau=N1-1]
					\ShowNiveaus[shift=1.0,length=2.0,niveau=N1-3]

					\ShowEa[label,label-side=left,label-pos=0.2]
					\ShowGain[label,offset=-40mm]

					\draw[above] (N1-1) ++ (1,0) node {\small \hspace{-40mm}reactants};
					\draw[above] (N1-3) ++ (1,0) node {\small products};

				\end{endiagram}
			}

		% end subsection


		\subsection{Maxwell-Boltzmann Distribution Curve}

			The other way that activation energy can be seen is through a Maxwell-Boltzmann distribution curve.

			\graphdiagram[1.0]{
				\begin{tikzpicture}
					\def\kB{1.3806488e-23}% boltzmann constant
					\def\temperature{298}% room temperature
					\def\Beta{1/(\kB*\temperature)}
					\def\amu{1.660538921e-27}% atomar mass unit in kg
					\def\mass{0.9*\amu}

					\begin{axis}[
						axis lines		= left,
						domain			= 0:6000,
						xlabel			= $E_{k}(x)$,
						ylabel			= $N(x)$,
						xtick			= \empty,
						ytick			= \empty,
						axis line style	= very thick,
						x label style	= {at={(0.4,-0.01)}},
						y label style	= {at={(0,0.5)}},
						ymin			= 0,
						ymax			= 0.0004
					]

						\addplot[color = black, name path = fn, very thick]{
							sqrt(2/pi)*(\mass*\Beta)^(3/2)*x^2*exp(-.5*\mass*\Beta*x^2)
						};

						\path[name path global = xaxis] (axis cs: 0,0) -- (axis cs: 6000,0);
						\node at (axis cs: 4000,-0.0000325) {$E_{a}$};

						\addplot[color = red, fill = blue, fill opacity = 0.50] fill between[
							of				= fn and xaxis,
							soft clip		= { domain = 4000:6000 },
						];

						\path[name path global = dtd] (axis cs:4000,0) -- (axis cs:4000,0.00020);

						\draw[
							ultra thick, red, dashed,
							name intersections = {of = fn and dtd, name = k},
							name intersections = {of = xaxis and dtd, name = m}
						] (k-1) -- (m-1);

					\end{axis}
				\end{tikzpicture}
			}

			The shaded area represents the total number of molecules whose energy is greater than the activation energy, and can undergo a
			successful reaction. Note that the reactant molecules must still be oriented correctly for a successful reaction.

			\pagebreak
			Activation energy is independent of the enthalpy change of the reaction, and instead depends on the reactants themselves; it is always
			positive, because bonds are always broken first before new bonds are formed.

			\begin{bulletlist}

			& If reactants are similarly charged, \ea{} will be high --- the reverse is also true
			& If the bonds that need to be broken are strong, \ea{} will be high

			\end{bulletlist}

			Activation energy also affects the rate of a reaction; the lower the \ea{}, the larger the fraction of molecules that are at or above
			this energy. Hence, the frequency of effective collisions increases, and rate of reaction increases as well.

		% end subsection
	% end section


	\section{Rates of Reaction}

		The rate of a reaction is primarily manifested through the rates of change of the concentrations of the reactants and products. That is,
		the faster the concentration of products increases, the faster the rate of reaction. Similarly, the faster the decrease of
		concentration of reactants, the faster the rate of reaction.

		\graphdiagram[1.0]{
			\begin{tikzpicture}

				\begin{axis}[
					axis lines		= left,
					domain			= 0:10,
					xlabel			= time/s,
					ylabel			= reactant conc.,
					axis line style	= very thick,
					height			= 80mm,
					enlarge y limits = 0.15
				]

					\addplot[color = black, name path = fn, very thick]{
						10 * e^(-0.25 * x)
					};

					\addplot[color = blue, very thick, restrict x to domain = 0:2]{
						-2.5 * x + 10
					};

					\addplot[color = red, very thick, restrict x to domain = 3:6]{
						-0.85 * x + 7.00
					};

					\node at (axis cs: 3,8) {initial rate};
					\node at (axis cs: 7,4) {instantaneous rate};

				\end{axis}

			\end{tikzpicture}
		}

		From a concentration-time graph, it is possible to determine the rate of a reaction through measuring the gradient.
		The initial rate of reaction is the gradient at $t = 0$, the rate of reaction at a given point in time is simply the gradient at that
		point, and the average rate of reaction is simply the rate of change of concentration across the graph.


		\pagebreak

		The rate equation outlines the relationship between the concentration of a given reactant and the actual rate of the reaction.
		All its components must be found experimentally, through a tedious process of actually conducting an experiment. It has this
		general form:

		\mathdiagram{
			\[ R = k[A]\sps{x}[B]\sps{y} \cdots [C]\sps{z} \]
		}

		$R$ is the rate of the reaction (in \si{\molarConc\per\second}). $k$ is the constant of proportionality
		(discussed below); $[A]$, $[B]$ etc. are the concentrations of each reactant, and $x$, $y$ etc. are the orders of reaction
		with respect to that reactant. For single-direction equations (ie. not dynamic equilibrium reactions), the rate equation typically
		contains only the reactants and not the products.

		The orders of reaction of each reactant are independent of the stoichiometric coefficients of those reactants in
		the actual equation. That is to say, an equation with \ch{4 CO2} does not mean that the order of reaction with respect to
		\ch{CO2} has to be 4. (it can be, but it's coincidental)



		\subsection{Order of Reaction}

			The order of a reaction with respect to a certain reactant determines the magnitude of the change in the rate of reaction when the
			concentration of that reactant is changed. Since the orders of the reaction are exponents, it can be inferred that the
			change in the rate of reaction would be equal to the change in the concentration of a given reactant, to the power
			of its order.

			For instance, given this reaction and its rate equation:

			\txtreactioneqntwo{
				\ch{2 NO \stG} + \ch{2 H2 \stG} \arrow \ch{N2 \stG} + \ch{2 H2O \stG}
			}{Rate = k[\ch{NO}]\sps{2}[\ch{H2}]}

			When the concentration of \ch{H2} is doubled, the rate of the reaction will double. However, when $[NO]$ is doubled, the rate of the
			reaction will quadruple. This also shows that the coefficients of the reactants in the chemical equation have nothing to do
			with the order of reaction.

			Orders of reaction are not restricted to nonzero integers, of course; if the order of reaction of a reactant is negative, then when
			its concentration is increased, the rate of reaction will decrease; fractional orders also exist.

			If the order of reaction of a reactant is 0, it is usually omitted from the rate equation, as changing its concentration has no
			effect on the rate of the reaction.

			The overall order of a reaction is simply the sum of the individual orders of its reactants.


		% end subsection


		\pagebreak
		\subsection{Rate Constant}

			The rate constant $k$ is an amalgamation of two factors; the activation energy of the reaction and the temperature
			of the system.

			In the presence of catalysts (discussed later), the activation energy of the reaction will decrease. As the concentrations of the
			reactants are not affected, the increase in the rate of reaction is manifested in the increase of the rate constant. Similarly, a
			change in temperature will also affect the rate constant, and hence the rate of the reaction.

			Since the concentration of the reactants is represented in \si{\molarConc} and the final rate of the reaction in \si{\molarConc\per\second},
			the rate constant must have appropriate units to \enquote{balance} the units of the rate equation; the units of the rate constant
			can be determined by simple dimensional analysis.


		% end subsection

	% end section




	\section{Determination of Reaction Order}

		\subsection{Initial Rates Method}

			A reactant’s order of reaction can be determined through the measurement of its initial rate, whilst changing the initial
			concentration of one or more reactants.

			Given the initial concentrations of all reactants and the initial rate (for two equations), a relationship between the rates can be
			determined. For example:

			\mathdiagram{
				\[ \frac{R_{1}}{R_{2}} = \frac{k(0.03)^{x}(1.0)^{y}(0.5)^{z}}{k(0.01)^{x}(1.0)^{y}(0.5)^{z}} \]
			}[Given that the initial rates $R_{1}$ and $R_{2}$ are known, the unknowns ($x$, $y$, and $z$) can be found.]
		% end subsubsection


		\subsection{Large Excess Method}

			The other method of determining the order of a reaction would be to use all but one reactant in large excess. For the purposes of this
			method, large excess is defined as a concentration 10x or more than the usual. This means that the concentrations of the other
			reactants appear to be constant, making them pseudo-zero-order.

			The reaction must be carried out once per reactant; an aliquot of the reaction mixture is first withdrawn, then the concentration of the
			reactant in question is measured, and then finally plotted in a concentration-time graph. A few things are in order:

			\begin{bulletlist}
				&	The withdrawn aliquot must be quenched, either by using ice (cooling), or removing the catalyst if the reaction cannot proceed
					without it.

				&	The aliquot is usually titrated with something to determine its concentration.
			\end{bulletlist}

		% end subsubsection

	% end subsection




	\pagebreak
	\section{Rate Graphs}

		\subsection{Concentration-Time Graph}

			A concentration-time graph is arguably the most versatile and useful graph. In such graphs, the rate of the reaction is represented
			by the gradient of the graph. As such, the order of the reaction can be determined by the shape of the graph.

			Concentration-time graphs are often used with the \enquote{large-excess} method, so that only the concentration of the reactant being
			observed appears to change.


			\subsubsection{Zeroth Order Reactions}

				For the zero-order reaction graph below, the rate of the reaction, which is the gradient, is constant even as $[E]$ decreases.
				Thus, the rate is unaffected by the concentration of $[E]$, and is zero order with respect to $[E]$.

				\graphdiagram[1.0]{
					\begin{tikzpicture}

						\begin{axis}[
							axis lines		= left,
							domain			= 0:8,
							xlabel			= time/s,
							ylabel			= reactant conc.,
							axis line style	= very thick,
							height			= 80mm,
							enlarge y limits = 0.15
						]

							\addplot[color = black, very thick]{
								-2.0 * x + 20
							};

						\end{axis}

					\end{tikzpicture}
				}[The gradient of the graph is $-kt$, and the y-intercept is the initial concentration, $[E]_{0}$.]

			% end subsubsection




			\pagebreak
			\subsubsection{First Order Reactions}

				For first-order reactions, the concentration of the reactant decreases exponentially with time, and it has a constant half-life.
				This graph is a curve, because the rate of the reaction decreases as $[A]$ decreases --- however, the rate of change of this
				decrease is constant.


				\graphdiagram[0.9]{
					\begin{tikzpicture}

						\begin{axis}[
							axis lines		= left,
							domain			= 0:1200,
							xlabel			= time/s,
							ylabel			= reactant conc.,
							axis line style	= very thick,
							height			= 80mm,
							width			= 120mm,
							ytick			= {1, 0.5, 0.25, 0.125, 0},
							yticklabel style= {/pgf/number format/.cd,fixed zerofill,precision=3}
						]

							\addplot[color = black, very thick]{
								e^(-0.0025 * x)
							};


							\addplot[color = red, very thick, restrict x to domain = 0:280]{0.5};
							\addplot[color = red, very thick] coordinates {(280, 0) (280, 0.5)};

							\addplot[color = red, very thick, restrict x to domain = 0:560]{0.25};
							\addplot[color = red, very thick] coordinates {(560, 0) (560, 0.25)};

							\addplot[color = red, very thick, restrict x to domain = 0:840]{0.125};
							\addplot[color = red, very thick] coordinates {(840, 0) (840, 0.125)};

							\addplot[<->] coordinates {(0, 0.035) (280, 0.035)};
							\addplot[<->] coordinates {(280, 0.035) (560, 0.035)};
							\addplot[<->] coordinates {(560, 0.035) (840, 0.035)};

							\node at (axis cs: 140, 0.07) {$t_{\sfrac{1}{2}}$};
							\node at (axis cs: 420, 0.07) {$t_{\sfrac{1}{2}}$};
							\node at (axis cs: 700, 0.07) {$t_{\sfrac{1}{2}}$};

						\end{axis}
					\end{tikzpicture}
				}[The half-life, $t_{\sfrac{1}{2}}$, is fixed.]


				The half-life of a reaction is defined as the time taken for the concentration of a reactant to decrease to half its previous value.
				As it can be seen, in the graph above the concentration decreases from 0.10 to 0.05 to 0.025 to 0.0125 and so on, every 280 seconds.

				Conversely, half-life can also be applied to the concentration of a product; it is the time taken for it to increase from 0 to 0.5,
				from 0.5 to 0.75, from 0.75 to 0.875 and so on. Note that the product concentration is just $1 - \mathrm{reactant\; concentration}$.


				\graphdiagram[0.9]{
					\begin{tikzpicture}

						\begin{axis}[
							axis lines		= left,
							domain			= 0:1200,
							xlabel			= time/s,
							ylabel			= product conc.,
							axis line style	= very thick,
							height			= 80mm,
							width			= 120mm,
							ytick			= {1, 0.875, 0.75, 0.5, 0},
							yticklabel style= {/pgf/number format/.cd,fixed zerofill,precision=3}
						]

							\addplot[color = black, very thick]{
								1 - e^(-0.0025 * x)
							};


							\addplot[color = red, very thick, restrict x to domain = 0:280]{0.5};
							\addplot[color = red, very thick] coordinates {(280, 0) (280, 0.5)};

							\addplot[color = red, very thick, restrict x to domain = 0:560]{0.75};
							\addplot[color = red, very thick] coordinates {(560, 0) (560, 0.75)};

							\addplot[color = red, very thick, restrict x to domain = 0:840]{0.875};
							\addplot[color = red, very thick] coordinates {(840, 0) (840, 0.875)};

							\addplot[<->] coordinates {(0, 0.035) (280, 0.035)};
							\addplot[<->] coordinates {(280, 0.035) (560, 0.035)};
							\addplot[<->] coordinates {(560, 0.035) (840, 0.035)};

							\node at (axis cs: 140, 0.07) {$t_{\sfrac{1}{2}}$};
							\node at (axis cs: 420, 0.07) {$t_{\sfrac{1}{2}}$};
							\node at (axis cs: 700, 0.07) {$t_{\sfrac{1}{2}}$};

						\end{axis}
					\end{tikzpicture}
				}


				\pagebreak
				For an overall first-order reaction, the half-life can be represented with this equation:

				\mathdiagram{
					\[ t_{\sfrac{1}{2}} = \frac{ln(2)}{k} \]
				}

				This means that the half-life is independent of the concentration of any reactant or product, and is only dependent on $k$,
				the rate constant.

			% end subsubsection


			\subsubsection{Second Order (and above) Reactions}

				The graph for second-order reactions has the same general exponential shape as the graph of a first-order reaction,
				except that the half-life is not constant; this is the key difference between a first-order reaction and
				reactions of second-order and above.


				\graphdiagram[1.0]{
					\begin{tikzpicture}

						\begin{axis}[
							axis lines		= left,
							domain			= 0:8000,
							xlabel			= time/s,
							ylabel			= reactant conc.,
							axis line style	= very thick,
							height			= 80mm,
							width			= 120mm,
							ytick			= {1, 0.5, 0.25, 0.125, 0},
							yticklabel style= {/pgf/number format/.cd,fixed zerofill,precision=3}
						]

							\addplot[color = black, very thick]{
								e^(-0.0025 * x^0.8)
							};


							\addplot[color = red, very thick, restrict x to domain = 0:1]{0.5};
							\addplot[color = red, very thick] coordinates {(1130, 0) (1130, 0.5)};

							\addplot[color = red, very thick, restrict x to domain = 0:2]{0.25};
							\addplot[color = red, very thick] coordinates {(2660, 0) (2660, 0.25)};

							\addplot[color = red, very thick, restrict x to domain = 0:3]{0.125};
							\addplot[color = red, very thick] coordinates {(4460, 0) (4460, 0.125)};

							\addplot[color = red, very thick, restrict x to domain = 0:3]{0.0625};
							\addplot[color = red, very thick] coordinates {(6400, 0) (6400, 0.0625)};

							\addplot[<->] coordinates {(0.00, 0.035) (1130, 0.035)};
							\addplot[<->] coordinates {(1130, 0.035) (2660, 0.035)};
							\addplot[<->] coordinates {(2660, 0.035) (4460, 0.035)};
							\addplot[<->] coordinates {(4460, 0.035) (6400, 0.035)};

							\node at (axis cs: 765, 0.07) {$t_{\sfrac{1}{2}}$};
							\node at (axis cs: 1895, 0.07) {$t_{\sfrac{1}{2}}$};
							\node at (axis cs: 3560, 0.07) {$t_{\sfrac{1}{2}}$};
							\node at (axis cs: 5430, 0.07) {$t_{\sfrac{1}{2}}$};

						\end{axis}

					\end{tikzpicture}
				}

				Here, the half-life increases as the concentration of A decreases, so this is not a first-order reaction. It is possible to
				determine the precise order of a reaction using a rate-concentration graph (below) if the half-life is not constant.

			% end subsubsection
		% end subsection




		\pagebreak
		\subsection{Rate-Concentration Graph}

			The other useful graph is a plot of rate of reaction against concentration of a given reactant. This graph can be
			obtained in two ways;

			\begin{bulletlist}
				& Measure the gradient at intervals along the concentration-time graph (this is the rate), plotting these against the
				concentration of the reactant at that time

				& Measuring various initial rates of reaction from changing the initial concentration of a reactant, and plotting those.
				Note that all but the zero-order graphs pass through the origin, as rate is 0 when $[A]$ is 0.
			\end{bulletlist}


			\subsubsection{Zeroth Order Reactions}

				If a reaction is zero-order with respect to a certain reactant, then the rate of reaction is independent of its concentration.
				Thus, since rate $= k[A]^{0}$, then $\mathrm{rate} = k$, and the result is a horizontal line. The rate is constant at all values of $[A]$.


				\graphdiagram[1.0]{
					\begin{tikzpicture}

						\begin{axis}[
							axis lines		= left,
							domain			= 0:10,
							xlabel			= {[A] / \si{\molarConc}},
							ylabel			= {rate / \si{\molarConc\per\second}},
							axis line style	= very thick,
							height			= 80mm,
							width			= 120mm,
							xtick			= \empty,
							ytick			= \empty,
						]

							\addplot[color = black, very thick]{
								10
							};

						\end{axis}

					\end{tikzpicture}
				}

			% end subsubsection

			\pagebreak
			\subsubsection{First Order Reactions}

				For first-order reactions, the rate increases with the concentration of the reactant in a linear manner.
				As such, rate $= k[A]$, which is in effect a $y = kx$ graph. Note that the graph must cut the origin, since the rate of
				reaction must be 0 when there is no reactant.

				\graphdiagram[1.0]{
					\begin{tikzpicture}

						\begin{axis}[
							axis lines		= left,
							domain			= 0:10,
							xlabel			= {[A] / \si{\molarConc}},
							ylabel			= {rate / \si{\molarConc\per\second}},
							axis line style	= very thick,
							height			= 80mm,
							width			= 120mm,
							xtick			= \empty,
							ytick			= \empty,
							ymax			= 10,
							xmax			= 8
						]

							\addplot[color = black, very thick, restrict y to domain = 0:8]{
								1.2 * x
							};

						\end{axis}

					\end{tikzpicture}
				}

			% end subsubsection

			\subsubsection{Second Order (and above) Reactions}

				For reactions that are second-order or higher, the graph simply looks like an increasing polynomial curve.

				\graphdiagram[1.0]{
					\begin{tikzpicture}

						\begin{axis}[
							axis lines		= left,
							domain			= 0:10,
							xlabel			= {[A] / \si{\molarConc}},
							ylabel			= {rate / \si{\molarConc\per\second}},
							axis line style	= very thick,
							height			= 80mm,
							width			= 100mm,
							xtick			= \empty,
							ytick			= \empty,
							ymax			= 10,
							xmax			= 8
						]

							\addplot[color = black, very thick, restrict y to domain = 0:8]{
								0.7 * x^2
							};

						\end{axis}

					\end{tikzpicture}
				}

				A better way to affirm the actual order of the reaction would be to plot a graph of rate against concentration
				squared --- resulting in a linear graph. A similar concept can be applied to the third order and above, plotting
				$[A]^{3}$ or $[A]^{4}$.

			% end subsubsection

		% end subsection

	% end section





	\section{Factors Affecting Rate of Reaction}

		Note that all the factors below with the \itl{exception} of concentration and pressure reflect the change in the rate of reaction
		through a change in the rate constant, $k$.

		\subsection{Concentration and Pressure}

			When the concentration or pressure (for gas) of a reactant is increased, there are \itl{more molecules of reactant
			in a given volume}; thus, they are closer together, and they collide more frequently. Thus, there will be a
			\itl{greater frequency of effective collisions}, resulting in an increase in the rate of reaction.

			However, there is a caveat --- there will only be an effect on the rate of reaction if the reactant in question appears
			in the rate equation, implying that it must be involved in the rate-determining step as well.

		% end subsection

		\subsection{Surface Area of Solids}

			If the reaction involves a solid reactant, then the exposed surface area of the block is also a determining factor. With a
			greater surface area, there is a \itl{greater probability} that other reactants can come into contact with the solid, thus
			\itl{increasing the effective frequency of collisions}, thus increasing the rate of reaction.

			As an example, a single large block of solid reactant will result in a slower rate of reaction as opposed to the use of a
			finely powdered solid reactant. This effects a change in $k$.

		% end subsection


		\subsection{Temperature}

			By increasing the temperature, the rate constant $k$ is also increased; there is a \itl{greater proportion} of molecules
			with kinetic energy greater than the activation energy. Hence, more of the collisions that occur will result in a successful reaction.

			\graphdiagram[1.0]{
				\begin{tikzpicture}

					\def\kB{1.3806488e-23}% boltzmann constant
					\def\to{300}% room temperature
					\def\tt{400}% room temperature
					\def\bo{1/(\kB*\to)}
					\def\bt{1/(\kB*\tt)}
					\def\amu{1.660538921e-27}% atomar mass unit in kg
					\def\mass{0.9*\amu}

					\begin{axis}[
						axis lines		= left,
						domain			= 0:6000,
						xlabel			= $E_{k}(x)$,
						ylabel			= $N(x)$,
						xtick			= \empty,
						ytick			= \empty,
						axis line style	= very thick,
						x label style	= {at={(0.4,-0.01)}},
						y label style	= {at={(0,0.5)}},
						ymin			= 0,
						ymax			= 0.0004
					]

						\addplot[color = black, name path = fn1, very thick]{
							sqrt(2/pi)*(\mass*\bo)^(3/2)*x^2*exp(-.5*\mass*\bo*x^2)
						};

						\addplot[color = black, name path = fn2, very thick]{
							sqrt(2/pi)*(\mass*\bt)^(3/2)*x^2*exp(-.5*\mass*\bt*x^2)
						};


						\path[name path global = xaxis] (axis cs: 0,0) -- (axis cs: 6000,0);
						\node at (axis cs: 4000,-0.0000325) {$E_{a}$};

						\addplot[color = red, fill = blue, fill opacity = 0.50] fill between[
							of				= fn2 and fn1,
							soft clip		= { domain = 4000:6000 },
						];

						\addplot[color = red, fill = green, fill opacity = 0.50] fill between[
							of				= fn1 and xaxis,
							soft clip		= { domain = 4000:6000 },
						];

						\node at (axis cs: 3000,0.00037) {\SI{300}{\kelvin}};
						\node at (axis cs: 4000,0.00027) {\SI{400}{\kelvin}};


						\path[name path global = dtd] (axis cs:4000,0) -- (axis cs:4000,0.00035);

						\draw[
							ultra thick, red, dashed,
							name intersections = {of = fn2 and dtd, name = k},
							name intersections = {of = xaxis and dtd, name = m}
						] (k-1) -- (m-1);

					\end{axis}

				\end{tikzpicture}
			}[When the temperature increases, the fraction of molecules with $\mathrm{KE} > E_{a}$ increases.]

		% end subsection


		\subsection{Effect of Light}

			Light can also have an effect. When certain molecules absorb light energy, it is manifested as kinetic energy, resulting in a
			greater proportion of molecules that are beyond the energy barrier of the activation energy.

		% end subsection



		\subsection{Catalysts}

			A catalyst provides an \itl{alternative pathway} to the original, without modifying the activation energy of the original
			reaction.

			The alternative pathway typically has a lower activation energy than the original reaction route, hence increasing the number
			of molecules that have a kinetic energy exceeding said activation energy.

			\graphdiagram[1.0]{
				\begin{tikzpicture}

					\def\kB{1.3806488e-23}% boltzmann constant
					\def\temp{300}% room temperature
					\def\beta{1/(\kB*\temp)}
					\def\amu{1.660538921e-27}% atomar mass unit in kg
					\def\mass{0.9*\amu}

					\begin{axis}[
						axis lines		= left,
						domain			= 0:6000,
						xlabel			= $E_{k}(x)$,
						ylabel			= $N(x)$,
						xtick			= \empty,
						ytick			= \empty,
						axis line style	= very thick,
						x label style	= {at={(0.4,-0.01)}},
						y label style	= {at={(0,0.5)}},
						ymin			= 0,
						ymax			= 0.0004
					]

						\addplot[color = black, name path = fn, very thick]{
							sqrt(2/pi)*(\mass*\beta)^(3/2)*x^2*exp(-.5*\mass*\beta*x^2)
						};

						\path[name path global = xaxis] (axis cs: 0,0) -- (axis cs: 6000,0);

						\node at (axis cs: 3500,-0.0000325) {$E_{a_{2}}$};
						\node at (axis cs: 4000,-0.0000325) {$E_{a_{1}}$};


						\addplot[color = red, fill = green, fill opacity = 0.50] fill between[
							of				= fn1 and xaxis,
							soft clip		= { domain = 3500:4000 },
						];

						\addplot[color = red, fill = blue, fill opacity = 0.50] fill between[
							of				= fn1 and xaxis,
							soft clip		= { domain = 4000:6000 },
						];

						\path[name path global = dtd1] (axis cs:3500,0) -- (axis cs:3500,0.00035);

						\draw[
							ultra thick, red, dashed,
							name intersections = {of = fn and dtd1, name = k},
							name intersections = {of = xaxis and dtd1, name = m}
						] (k-1) -- (m-1);


						\path[name path global = dtd2] (axis cs:4000,0) -- (axis cs:4000,0.00035);

						\draw[
							ultra thick, red, dashed,
							name intersections = {of = fn and dtd2, name = k},
							name intersections = {of = xaxis and dtd2, name = m}
						] (k-1) -- (m-1);


					\end{axis}

				\end{tikzpicture}
			}[The activation energy decreases from $E_{a_{1}}$ to $E_{a_{2}}$, increasing the fraction of molecules with $\mathrm{KE} > E_{a}$.]


			This effect can also be similarly illustrated with a energy profile diagram:

			\graphdiagram[1.0]{
				\begin{endiagram}[scale=1.5,offset=1.5,x-label-text=reaction coordinate,y-label-text=energy]
					\ENcurve[step=1.5]{0,4.5,2}
					\ENcurve[step=1.5,tikz={red,dotted}]{0,3,2}

					\ShowNiveaus[shift=-1.0,length=2.0,niveau=N1-1]
					\ShowNiveaus[shift=1.0,length=2.0,niveau=N1-3]

					\ShowEa[label,from={(N1-1) to (N1-2)}]
					\ShowGain[label]

					\draw[above] (N1-1) ++ (1,0) node {\small \hspace{-40mm}X};
					\draw[above] (N1-3) ++ (1,0) node {\small Y};

				\end{endiagram}
			}[The dotted line represents the alternative pathway that is taken, with a lower activation energy.]


		% end subsection

	% end section

% end part

























